Interactive Filter and Display of Hillary Clinton's Emails: \\A Cautionary Tale of Metadata

### 2017-04-11 VAST_submission/vgtc_journal_latex/paper.tex

On March 16, 2016 Wikileaks  published a searchable archive \cite{Wikileaks} containing the contents of more than 30,000 emails (and attachments) that were sent to and from Secretary Clinton on her private server.  The documents were provided as pdfs by the U.S. Department of State in response to Freedom of Information Act (FOIA) \cite{FOIA} requests.  The State Dept. also provided a searchable web archive of the documents, released in several instalments from May 2015 to March 2017\cite{StateDeptFOIA}.   Both sites provide a  useful tool for anyone searching for particular terms in these documents.  The Wikileaks site was put to much use by investigative reporters, and others, to search for topical news items.

What is critically missing from either site is a facility to learn summary, or statistical, features across all, or groups of,  emails.   A separate site where visual displays of summary features of the emails, which can be filtered by user selection, would go a long way towards filling this need.  The site would be \emph{complementary} to the more typical ``content search and record display'' site as provided by either online archive.   Add general internet search in a third browser, and the three combined would provide the means to actively explore the contents of any email archive.    One imagines, for example, having three browsers open simultaneously, one for the data base search (e.g. Wikileaks), one for statistical visualization and exploration (proposed here), and one for internet search (to provide important context for the other two services).  Each fills a need not met by the other two; the synergy of all three provide the user a powerful set of investigative tools.

In this paper, we describe an implementation of such a visual analytic service and illustrate it on the Secretary Clinton's e-mail archive.  The implementation is located at  \texttt{rshiny.math.uwaterloo.ca/clinton} where any visitor may interactively explore summary features  of the entire corpus of emails.   Different visual displays present different salient features from all selected emails; emails can be selected via a variety of interactive filters.   All displays are reactive and update simultaneously in response to user interaction with the data filters.  Together, filters and displays provide a visual analysis service that can be used to quickly learn more about, and to discover some (possibly unanticipated) patterns in, the emails.

The paper is organized as follows.  We provide a brief timeline on the private email server to provide sufficient context for the work.  Readers more familiar with the background may wish to skip to the next three sections where the implementation is described.  Section \ref{sect:data} discusses the source of the data, its extraction and cleaning,  details of the metadata,  redaction codes as quasi-metadata,  and our limited use of the actual email content. 
Section \ref{sect:Displays} describes the displays used and illustrates them on the data. These include a spoked network plot, an email volume time series plot, a scatterplot of daily sending times, a barplot of exemption codes, and term frequency displays. Section \ref{sect:Filters} presents the interactive filters that can be applied to change the displays.  With the implementation described, some interactive analyses of the Clinton email archive are conducted to give a quick overview of the utility of the service.  We find some interesting, and potentially contentious, patterns in the email correspondence and check these against information found from Google searches for the dates in which the patterns occur.  The section closes with caveats on over interpretation of these analyses.  Some closing thoughts are given in the final section.

@@

Nearly one year later, July 5, 2016,  FBI Director James Comey recommended that no charges be laid against Clinton for her use of a private email server \cite{nochargeFBI}.  On October 28, 2016, Comey revealed that, in a separate investigation into former Congressman Anthony Weiner, emails belonging to his wife Huma Abedin had been found on his laptop.  Since Abedin was a close aid to former Secretary Clinton, FBI investigations were reopened into the private server usage but closed again by Comey on November 6, 2016 without charges being laid \cite{nochargeFBINov}.  In the first case Comey and the FBI were criticized by republican pundits and in the second case by democratic pundits.

@@

The HTML was constructed from email printed out, redacted, and then provided as pdfs by the State Department.  Consequently, detailed email header information as would normally be available electronically is mostly missing.   All time stamps appear to be the local date and time at which the server sent or received that email (e.g. no time zone or other source time or IP chain information is available).  Moreover, because of redaction (typically FOIA exemption B6 ~\cite{FOIA}),  sender and receiver emails may have truncated domains,  contain only the person's name, or be missing altogether.  In cases where both \texttt{From} and \texttt{To} are entirely missing we do not impute values  (e.g. see \texttt{https://wikileaks.org/clinton-emails/emailid/31599}).

@@

\begin{figure}[h]
\begin{center}
\includegraphics[width=0.95\linewidth]{SpiralNetworkFullTime}
\caption{Inner Circle over all correspondence}
\label{fig:InnerCircleAll}
\end{center}
\end{figure}
Figure~\ref{fig:InnerCircleAll} shows the top 20 correspondents (or ``Inner Circle'')  over all emails in the collection.
Note that Huma Abedin appears both as a blue  \texttt{.gov} account and as a red non-government (\texttt{clintonemail.com}) account.   The other red account is that of Sidney Blumenthal.  This has been a source of some controversy  \cite{BlumenthalControversy, NYTBlumenthalBenghazi}, since he had been rejected by the State Department yet appears to have provided Secretary Clinton advice throughout her tenure.  Wendy Sherman appears as orange. Sherman was appointed by Clinton to be Under Secretary of State for Political Affairs in 2011 and the high frequency of Sherman emails dates from this time.  Nevertheless, from this archive the email address could not be identified definitively as being either government or non-government.  Either the address was redacted for privacy reasons (a B6 FOIA exemption), or it was simply unavailable in the header provided by the WikiLeaks HTML \cite{WikileaksPull} (e.g. occasionally, a blank header appeared in the WikiLeaks HTML that was not present in the pdf; we did not correct for these manually).

The width of each spoke is proportional to the volume of emails between correspondents.  Whenever the difference in volume is great enough,  correspondents are separated into two groups: those with the greatest correspondence have shorter spokes visually placing them ``closer'' to Clinton.  As Figure~\ref{fig:InnerCircleAll} shows shows, Clinton's closest inner circle are her closest aides Huma Abedin, Cheryl Mills, and Jake Sullivan.
\subsection{Email volume}
\label{sect:Displays:volume}
Figure~\ref{fig:VolumeAll} 
\begin{figure}[h]
\begin{center}
\includegraphics[width=0.95\linewidth]{VolumeFullTime}
\caption{Email Volume: All emails (top) ; Those sent by Clinton (bottom)}
\label{fig:VolumeAll}
\end{center}
\end{figure}
presents two time series showing the volume of emails for each day.  In grey at top is the daily total; in magenta at bottom is the total for the subset selected by the filters.  In Figure~\ref{fig:VolumeAll}, the bottom series is the total daily emails that were sent \emph{from} Clinton.

Over the four year scale shown in Figure~\ref{fig:VolumeAll}, the display is quite busy.  Even so, a few features stand out.  For example, there is a notable lack of email from the first few months of Clinton's tenure;  there is essentially no email from Clinton at the beginning and very little to her.

@@

\subsection{Email times}
\label{sect:Displays:times}
Figure~\ref{fig:TimesAll}  
\begin{figure}[h]
\begin{center}
\includegraphics[width=0.95\linewidth]{DailyFullTimeFromClinton}
\caption{Email time patterns: all emails sent from Clinton}
\label{fig:TimesAll}
\end{center}
\end{figure}
shows all emails sent by Clinton over the four year period.  Each email appears as a point determined by the time that it was sent.  The calendar date determines the position on the horizontal axis,  the (24hr) time of day determines the vertical.   Points are coloured red if any part of that email was redacted, and blue if it was released in full.   Alpha blending is used to minimize the effect of over-plotting.

@@

Figure \ref{fig:FOIAcodes}
\begin{figure}[h]
\begin{center}
\includegraphics[width=0.95\linewidth]{ExemptionsFullTime}
\caption{FOIA barplot: Exemption codes from all email}
\label{fig:FOIAcodes}
\end{center}
\end{figure}
is a barplot of all FOIA exemption codes found in all of the emails.  As can be seen, B1 redactions made for national security and foreign policy reasons account for a small portion of all redactions made,  though this amounts to more than 2,000 emails in the collection.   Nearly half of the emails contain B6 exemptions, those made to protect personal privacy.   The inter-agency, intra-agency exemption B5 is used in more than 8,000 emails.  This code has come under some criticism from advocates for transparent government as being overly applied \cite{unredactedB5}.  The remaining codes appear in very few emails and about 12,000 emails, or about a third of them, are not redacted at all.

@@

\subsection{Term frequency and TFIDF}
\label{sect:Displays:TFIDF}
The 20 terms (stemmed words excluding stopwords) which appear in the greatest number of the emails selected are displayed as those having ``Highest Frequency''.  
\begin{figure}[h]
\begin{center}
\begin{tabular}{ccc}
\includegraphics[width=0.4\linewidth]{HighFreqAll} &
$~~~~$ &
\includegraphics[width=0.4\linewidth]{HighTFIDFAll} 
\end{tabular}
\caption{Top 20 terms appearing in all emails}
\label{fig:top20all}
\end{center}
\end{figure}
Those terms which appear frequently within some emails but less frequently across emails will have a high tf-idf (term frequency - inverse document frequency) score.  The top 20 scoring of these in the selected emails are shown in the ``TFIDF'' display. 
The top 20 shown in Figure~\ref{fig:top20all} are for all emails in the corpus; these will change depending on the filtering.

The terms give some limited insight into the contents of the email.  As seen in Figure~\ref{fig:top20all}, for example, the first names of two of Clinton's closest aides, Jake Sullivan and Cheryl Mills,  appear under TFIDF but not  as high frequency terms.  One problem with tf-idf for this e-mail corpus is that many emails contain email chains which grow as each person replies; this could magnify the within email frequency for some terms.

As mentioned in Section~\ref{sect:data:contentinfo}, the stemming and stopword removal provided by the package \texttt{tm} in \texttt{R} is challenged by this messy and non-standard data. The list of stopwords had to be supplemented so as to avoid uninformative action verbs such as 'will' and 'can'. The stemming algorithm was also challenged by typographical errors in emails and the proliferation of acronyms (e.g. for individuals and government abbreviations).

@@

\subsection{FOIA exemptions}
Multiple filtering by FOIA exemption codes is supported using a box selection area. 
\begin{figure}[h]
\begin{center}
\includegraphics[width=0.55\linewidth]{FOIABoxImage}
\caption{FOIA codes:  Emails containing any of these codes appear}
\label{fig:selectFOIA}
\end{center}
\end{figure}
Figure~\ref{fig:selectFOIA} displays the codes that are selected for email display.  Any email having one or more of these codes will appear in the displays.  To not have emails display with a B1 exemption code, for example, then the user selects and deletes the B1 from the list (multiple selection is enabled).  Or to have only the emails which  contain B1 redactions displayed, all other codes can deleted from the list.

@@

In this section, we explore a few avenues of enquiry that might naturally arise.  At best these only hint at the analytic power that interactive filtering and display can bring to bear on an email archive.  The user is encouraged to try it themselves at \texttt{rshiny.math.uwaterloo.ca/clinton}to get a fuller appreciation of the power of the tool.  Those who are exploring the corpus will find it useful for generating hypotheses that might be tested against other sources.   This is the approach taken here.  Conversely, those more knowledgeable about the contexts surrounding Secretary Clinton and her private email server might use the visual analytic tool more as a means to test hypotheses already held.

@@

Figure~\ref{fig:lastTwoYears} 
\begin{figure}[h]
\begin{center}
\includegraphics[width=0.95\linewidth]{EmailVolumeB1LastTwoYears}
\caption{Last two years:  Selected Emails are those that are B1 redacted}
\label{fig:lastTwoYears}
\end{center}
\end{figure}
shows the email volume time series over this period. 
This is accomplished by simply having the right end of the time slider of Figure~\ref{fig:timeSlider} as far right as possible and then moving the left most end towards the middle of the time period until January 1, 2011 is reached.   Note, however, that in addition to filtering on time we have chosen to also filter on FOIA exemption code, selecting only those emails containing one or more B1 (national security and foreign policy matters) redactions.  The magenta time series at the bottom then shows all emails, sent or received by Clinton, in the last two years of her tenure which were redacted in part by the State Department (after the fact, not at the time) as being B1 FOIA exempt.  As can be seen, there are \emph{relatively} few of these.

Recall from Section~\ref{sect:Displays:circle} that the tallest spike of Figure~\ref{fig:lastTwoYears} corresponds to the beginning of the battle for Tripoli (August 21, 2011).   The next largest spikes appear at the far left of this time window, at the beginning of 2011.  We might choose to compress the time window more, to focus on the early 2011 section of the emails.  Bringing the right end of the time slider in to April 15, 2011 produces the series of Figure~\ref{fig:early2011B1}.
\begin{figure}[h]
\begin{center}
\includegraphics[width=0.95\linewidth]{EmailVolumeB1JanApril2011}
\caption{Jan. 1 -- Apr 15, 2011:  Selected Emails are those that are B1 redacted}
\label{fig:early2011B1}
\end{center}
\end{figure}

As can be seen, there continue to be highs and lows of email activity.   The highest peak now is at the left, though closer to the centre of Figure~\ref{fig:early2011B1} than it had been in Figure~\ref{fig:lastTwoYears} .  This is January 29, 2011, the day after President Mubarak of Egypt has ordered the army into the streets of Cairo to quell protests \cite{cairoTanks2011, cairoResponse2011}.  Dipping into the contents of the Wikileaks email archive, much of the email on this day is seen to be about managing this Egyptian crisis.

@@

The readjusted time series are shown in Figure~\ref{fig:EmailVolumeB1LibyaBuildup}.
\begin{figure}[h]
\begin{center}
\includegraphics[width=0.95\linewidth]{EmailVolumeB1LibyaBuildup}
\caption{Feb. 25 -- March 15, 2011:  Selected Emails are those that are B1 redacted}
\label{fig:EmailVolumeB1LibyaBuildup}
\end{center}
\end{figure}
The left most spike in Figure~\ref{fig:EmailVolumeB1LibyaBuildup}  is the rightmost spike of the pair of approximately 80 emails from Figure~\ref{fig:early2011B1}.   
Of course all other displays have been updating every time the time slider is adjusted.

Some sense of the content of the emails can be had from the updated term frequency displays shown in Figure~\ref{fig:TermFreqB1LibyaBuildup}.
\begin{figure}[h]
\begin{center}
\begin{tabular}{ccc}
\includegraphics[width=0.4\linewidth]{TFIDFB1LibyaBuildup} &
$~~~~$ &
\includegraphics[width=0.4\linewidth]{TermFreqB1LibyaBuildup} 
\end{tabular}
\caption{Frequent terms: B1 only emails, Feb. 25 -- March 18, 2011}
\label{fig:TermFreqB1LibyaBuildup}
\end{center}
\end{figure}
Not surprisingly,  Libya and its leader figure prominently in this restricted set of emails.  
Perhaps more surprising is a term like ``windrush''.  A little investigation shows that this is a reference to ``Windrush Ventures'' (e.g. see  \cite{windrushTelegraph, windrushGuardian}),  a company owned by former U.K. Prime Minister Tony Blair .  ``Windrush'' appears at the end of emails as part of Mr. Blair's electronic signature.   
Mr. Blair is part of this B1 selected correspondence.

Other restrictions, as measured by the FOIA exemption codes, can be seen in Figure~\ref{fig:FOIAB12011LibyaBuildup}.
\begin{figure}[h]
\begin{center}
\includegraphics[width=0.95\linewidth]{FOIAB12011LibyaBuildup}
\caption{FOIA code distribution: All  B1 redacted, Feb. 25 -- March 18, 2011}
\label{fig:FOIAB12011LibyaBuildup}
\end{center}
\end{figure}
There appear to be about 100 emails in the selected set.  In addition to having B1 redactions, many of them also have B5 (inter- or intra-agency) and B6 (personal) redactions.

We might now consider the composition of the inner circle for this set of emails.  Figure~\ref{fig:InnerCircle2011B1LibyaBuildup} 
\begin{figure}[h]
\begin{center}
\includegraphics[width=0.95\linewidth]{InnerCircle2011B1LibyaBuildup}
\caption{Inner Circle: All  B1 redacted, Jan. 1 -- Apr 15, 2011}
\label{fig:InnerCircle2011B1LibyaBuildup}
\end{center}
\end{figure}
shows the set of correspondents.  Not surprisingly, Jake Sullivan is the principal correspondent.  Huma Abedin is next, though she is using two emails, one a government account the other a personal \texttt{clintonemail.com} account.   Sidney Blumenthal, a non-government employee, is in this inner circle of B1 exempted emails.  So too is former U.K. Prime Minister Tony Blair; this is the unidentifiable domain account \texttt{aclb}.
It should be noted that there could be other accounts which do not appear in the inner circle because their emails were not present in the email record.  See Section~\ref{sect:data:metadata}.

We could continue to drill down into this information, checking the (unredacted) contents of the emails themselves from either the Wikileaks or State Department sources.  We might also more closely connect email dates and correspondents with world events.   Instead, we now turn our attention to another feature which appears in Figure~\ref{fig:lastTwoYears}.

\subsection{An email gap?}
\label{sect:emailGap}
Towards the very right of Figure~\ref{fig:lastTwoYears} there appears to be an anomaly in both time series.  There is a noticeable flat area at zero in both series.

If we move the leftmost time sliders to the June 1, 2012, and the rightmost to the end, February 1, 2013, we have the email volumes for the last 8 months of the archive.  The resulting series are shown in Figure~\ref{fig:EmailVolumeB1Jun1On}.
\begin{figure}[h]
\begin{center}
\includegraphics[width=0.95\linewidth]{EmailVolumeB1Jun1On}
\caption{June 1, 2012 -- February 1, 2013:  Selected Emails are those that are B1 redacted}
\label{fig:EmailVolumeB1Jun1On}
\end{center}
\end{figure}
There appears to be no emails for a fairly large chunk of time.  In fact, there are few or no emails in the archive from October 30 to November 9, 2012.  
This seems curious, given that the archive is supposed to be complete in Secretary Clinton's work-related email during this time period.

The block of no email appears as an even more unusual pattern in Figure~\ref{fig:teaser}(a).  There the block is conspicuous by the absence amongst the emails from September 11 to November 23, 2012.  The email on which the displays of Figure~\ref{fig:teaser} are based, is no longer filtered by FOIA code.  It contains all email in the archive in this time period.  Direct examination of the archives at Wikileaks and the State Department show no mail in this empty block beyond that seen in Figure~\ref{fig:teaser}(a).

>>>>

On March 16, 2016 Wikileaks published a \href{https://wikileaks.org/clinton-emails/}{\color{blue}searchable archive} \cite{Wikileaks} containing the contents of more than 30,000 emails (and attachments) that were sent to and from Secretary Clinton on her private server.  The documents were provided as pdfs by the U.S. Department of State in response to Freedom of Information Act (FOIA) \cite{FOIA} requests.  The State Dept. also provided a \href{https://foia.state.gov/Search/Results.aspx?collection=Clinton_Email}{\color{blue}searchable web archive} of the documents, released in several instalments from May 2015 to March 2017\cite{StateDeptFOIA}.   Both sites provide a  useful tool for anyone searching for particular terms in these documents.  The Wikileaks site was put to much use by investigative reporters, and others, to search for topical news items.

What is critically missing from either site is a facility to learn summary, or statistical, features across all, or groups of,  emails.   A separate site where visual displays of summary features of the emails, which can be filtered by user selection, would go a long way towards filling this need.  The site would be \emph{complementary} to the more typical ``content search and record display'' site as provided by either online archive.   Add general internet search in a third browser, and the three combined would provide the means to actively explore the contents of any email archive.    One imagines, for example, having three browsers open simultaneously, one for the data base search (e.g. \href{https://wikileaks.org/clinton-emails/}{\color{blue}WikiLeaks}), one for statistical visualization and exploration (\href{https://rshiny.math.uwaterloo.ca/clinton}{\color{blue} proposed here}), and one for \href{https://google.com}{\color {blue} internet search} (to provide important context for the other two services).  Each fills a need not met by the other two; the synergy of all three provide the user a powerful set of investigative tools.

In this paper, we describe an implementation of such a visual analytic service and illustrate it on the Secretary Clinton's e-mail archive.  The implementation is located at
\href{https://rshiny.math.uwaterloo.ca/clinton}{\color{blue} \texttt{rshiny.math.uwaterloo.ca/clinton} }
where any visitor may interactively explore summary features  of the entire corpus of emails.   Different visual displays present different salient features from all selected emails; emails can be selected via a variety of interactive filters.   All displays are reactive and update simultaneously in response to user interaction with the data filters.  Together, filters and displays provide a visual analysis service that can be used to quickly learn more about, and to discover some (possibly unanticipated) patterns in, the emails.

The paper is organized as follows.  We provide a brief timeline on the private email server to provide sufficient context for the work.  Readers more familiar with the background may wish to skip to the next three sections where the implementation is described.   \autoref{sect:data} discusses the source of the data, its extraction and cleaning,  details of the metadata,  redaction codes as quasi-metadata,  and our limited use of the actual email content. 
Section \autoref{sect:Displays} describes the displays used and illustrates them on the data. These include a spoked network plot, an email volume time series plot, a scatterplot of daily sending times, a barplot of exemption codes, and term frequency displays. Section \autoref{sect:Filters} presents the interactive filters that can be applied to change the displays.  With the implementation described, some interactive analyses of the Clinton email archive are conducted to give a quick overview of the utility of the service.  We find some interesting, and potentially contentious, patterns in the email correspondence and check these against information found from Google searches for the dates in which the patterns occur.  The section closes with caveats on over interpretation of these analyses.  Some closing thoughts are given in the final section.

@@

Nearly one year later, July 5, 2016,  FBI Director James Comey recommended that no charges be laid against Clinton for her use of a private email server \cite{nochargeFBI}.  On October 28, 2016, Comey revealed that, in a separate investigation into former Congressman Anthony Weiner, emails belonging to his wife Huma Abedin had been found on his laptop.  Since Abedin was a close aid to former Secretary Clinton, FBI investigations were reopened into the private server usage but closed again by Comey on November 6, 2016 without charges being laid \cite{nochargeFBINov}.  In the first case Comey and the FBI were criticized by Republican pundits and in the second case by Democrat pundits.

@@

The HTML was constructed from email printed out, redacted, and then provided as pdfs by the State Department.  Consequently, detailed email header information as would normally be available electronically is mostly missing.   All time stamps appear to be the local date and time at which the server sent or received that email (e.g. no time zone or other source time or IP chain information is available).  Moreover, because of redaction (typically FOIA exemption B6 ~\cite{FOIA}),  sender and receiver emails may have truncated domains,  contain only the person's name, or be missing altogether.  In cases where both \texttt{From} and \texttt{To} are entirely missing we do not impute values  (e.g. see \href{https://wikileaks.org/clinton-emails/emailid/31599}{\color {blue} \texttt{https://wikileaks.org/clinton-emails/emailid/31599}}).

@@

\begin{figure}[h]
\begin{center}
\includegraphics[width=0.95\linewidth]{SpiralNetworkFullTime}
\caption{Inner Circle over all correspondence}
\label{fig:InnerCircleAll}
\end{center}
\end{figure}
\autoref{fig:InnerCircleAll} shows the top 20 correspondents (or ``Inner Circle'')  over all emails in the collection.
Note that Huma Abedin appears both as a blue  \texttt{.gov} account and as a red non-government (\texttt{clintonemail.com}) account.   The other red account is that of Sidney Blumenthal.  This has been a source of some controversy  \cite{BlumenthalControversy, NYTBlumenthalBenghazi}, since he had been rejected by the State Department yet appears to have provided Secretary Clinton advice throughout her tenure.  Wendy Sherman appears as orange. Sherman was appointed by Clinton to be Under Secretary of State for Political Affairs in 2011 and the high frequency of Sherman emails dates from this time.  Nevertheless, from this archive the email address could not be identified definitively as being either government or non-government.  Either the address was redacted for privacy reasons (a B6 FOIA exemption), or it was simply unavailable in the header provided by the WikiLeaks HTML \cite{WikileaksPull} (e.g. occasionally, a blank header appeared in the WikiLeaks HTML that was not present in the pdf; we did not correct for these manually).

The width of each spoke is proportional to the volume of emails between correspondents.  Whenever the difference in volume is great enough,  correspondents are separated into two groups: those with the greatest correspondence have shorter spokes visually placing them ``closer'' to Clinton.  As \autoref{fig:InnerCircleAll} shows shows, Clinton's closest inner circle are her closest aides Huma Abedin, Cheryl Mills, and Jake Sullivan.
\subsection{Email volume}
\label{sect:Displays:volume}
\autoref{fig:VolumeAll} 
\begin{figure}[h]
\begin{center}
\includegraphics[width=0.95\linewidth]{VolumeFullTime}
\caption{Email Volume: All emails (top) ; Those sent by Clinton (bottom)}
\label{fig:VolumeAll}
\end{center}
\end{figure}
presents two time series showing the volume of emails for each day.  In grey at top is the daily total; in magenta at bottom is the total for the subset selected by the filters.  In \autoref{fig:VolumeAll}, the bottom series is the total daily emails that were sent \emph{from} Clinton.

Over the four year scale shown in \autoref{fig:VolumeAll}, the display is quite busy.  Even so, a few features stand out.  For example, there is a notable lack of email from the first few months of Clinton's tenure;  there is essentially no email from Clinton at the beginning and very little to her.

@@

\subsection{Email times}
\label{sect:Displays:times}
\autoref{fig:TimesAll}  
\begin{figure}[h]
\begin{center}
\includegraphics[width=0.95\linewidth]{DailyFullTimeFromClinton}
\caption{Email time patterns: all emails sent from Clinton}
\label{fig:TimesAll}
\end{center}
\end{figure}
shows all emails sent by Clinton over the four year period.  Each email appears as a point determined by the time that it was sent.  The calendar date determines the position on the horizontal axis,  the (24hr) time of day determines the vertical.   Points are coloured red if any part of that email was redacted, and blue if it was released in full.   Alpha blending is used to minimize the effect of over-plotting.

@@

\autoref{fig:FOIAcodes}
\begin{figure}[h]
\begin{center}
\includegraphics[width=0.95\linewidth]{ExemptionsFullTime}
\caption{FOIA barplot: Exemption codes from all email}
\label{fig:FOIAcodes}
\end{center}
\end{figure}
is a barplot of all FOIA exemption codes found in all of the emails.  As can be seen, B1 redactions made for national security and foreign policy reasons account for a small portion of all redactions made,  though this amounts to more than 2,000 emails in the collection.   Nearly half of the emails contain B6 exemptions, those made to protect personal privacy.   The inter-agency, intra-agency exemption B5 is used in more than 8,000 emails.  This code has come under some criticism from advocates for transparent government as being overly applied \cite{unredactedB5}.  The remaining codes appear in very few emails and about 12,000 emails, or about a third of them, are not redacted at all.

@@

\subsection{Term frequency and TFIDF}
\label{sect:Displays:TFIDF}
The 20 terms (stemmed words excluding stopwords) which appear in the greatest number of the emails selected are displayed as those having ``Highest Frequency''.  
\begin{figure}[h]
\begin{center}
\begin{tabular}{ccc}
\includegraphics[width=0.4\linewidth]{HighFreqAll} &
$~~~~$ &
\includegraphics[width=0.4\linewidth]{HighTFIDFAll} 
\end{tabular}
\caption{Top 20 terms appearing in all emails}
\label{fig:top20all}
\end{center}
\end{figure}
Those terms which appear frequently within some emails but less frequently across emails will have a high tf-idf (term frequency - inverse document frequency) score.  The top 20 scoring of these in the selected emails are shown in the ``TFIDF'' display. 
The top 20 shown in \autoref{fig:top20all} are for all emails in the corpus; these will change depending on the filtering.

The terms give some limited insight into the contents of the email.  As seen in \autoref{fig:top20all}, for example, the first names of two of Clinton's closest aides, Jake Sullivan and Cheryl Mills,  appear under TFIDF but not  as high frequency terms.  One problem with tf-idf for this e-mail corpus is that many emails contain email chains which grow as each person replies; this could magnify the within email frequency for some terms.

As mentioned in \autoref{sect:data:contentinfo}, the stemming and stopword removal provided by the package \texttt{tm} in \texttt{R} is challenged by this messy and non-standard data. The list of stopwords had to be supplemented so as to avoid uninformative action verbs such as 'will' and 'can'. The stemming algorithm was also challenged by typographical errors in emails and the proliferation of acronyms (e.g. for individuals and government abbreviations).

@@

\subsection{FOIA exemptions}
Multiple filtering by FOIA exemption codes is supported using a box selection area. 
\begin{figure}[h]
\begin{center}
\includegraphics[width=0.55\linewidth]{FOIABoxImage}
\caption{FOIA codes:  Emails containing any of these codes appear}
\label{fig:selectFOIA}
\end{center}
\end{figure}
\autoref{fig:selectFOIA} displays the codes that are selected for email display.  Any email having one or more of these codes will appear in the displays.  To not have emails display with a B1 exemption code, for example, then the user selects and deletes the B1 from the list (multiple selection is enabled).  Or to have only the emails which  contain B1 redactions displayed, all other codes can deleted from the list.

@@

In this section, we explore a few avenues of enquiry that might naturally arise.  At best these only hint at the analytic power that interactive filtering and display can bring to bear on an email archive.  The user is encouraged to try it themselves at \href{https://rshiny.math.uwaterloo.ca/clinton}{\color{blue} \texttt{rshiny.math.uwaterloo.ca/clinton} } to get a fuller appreciation of the power of the tool.  Those who are exploring the corpus will find it useful for generating hypotheses that might be tested against other sources.   This is the approach taken here.  Conversely, those more knowledgeable about the contexts surrounding Secretary Clinton and her private email server might use the visual analytic tool more as a means to test hypotheses already held.

@@

\autoref{fig:lastTwoYears} 
\begin{figure}[h]
\begin{center}
\includegraphics[width=0.95\linewidth]{EmailVolumeB1LastTwoYears}
\caption{Last two years:  Selected Emails are those that are B1 redacted}
\label{fig:lastTwoYears}
\end{center}
\end{figure}
shows the email volume time series over this period. 
This is accomplished by simply having the right end of the time slider of \autoref{fig:timeSlider} as far right as possible and then moving the left most end towards the middle of the time period until January 1, 2011 is reached.   Note, however, that in addition to filtering on time we have chosen to also filter on FOIA exemption code, selecting only those emails containing one or more B1 (national security and foreign policy matters) redactions.  The magenta time series at the bottom then shows all emails, sent or received by Clinton, in the last two years of her tenure which were redacted in part by the State Department (after the fact, not at the time) as being B1 FOIA exempt.  As can be seen, there are \emph{relatively} few of these.

Recall from \autoref{sect:Displays:circle} that the tallest spike of \autoref{fig:lastTwoYears} corresponds to the beginning of the battle for Tripoli (August 21, 2011).   The next largest spikes appear at the far left of this time window, at the beginning of 2011.  We might choose to compress the time window more, to focus on the early 2011 section of the emails.  Bringing the right end of the time slider in to April 15, 2011 produces the series of \autoref{fig:early2011B1}.
\begin{figure}[h]
\begin{center}
\includegraphics[width=0.95\linewidth]{EmailVolumeB1JanApril2011}
\caption{Jan. 1 -- Apr 15, 2011:  Selected Emails are those that are B1 redacted}
\label{fig:early2011B1}
\end{center}
\end{figure}

As can be seen, there continue to be highs and lows of email activity.   The highest peak now is at the left, though closer to the centre of \autoref{fig:early2011B1} than it had been in \autoref{fig:lastTwoYears} .  This is January 29, 2011, the day after President Mubarak of Egypt has ordered the army into the streets of Cairo to quell protests \cite{cairoTanks2011, cairoResponse2011}.  Dipping into the contents of the Wikileaks email archive, much of the email on this day is seen to be about managing this Egyptian crisis.

@@

The readjusted time series are shown in \autoref{fig:EmailVolumeB1LibyaBuildup}.
\begin{figure}[h]
\begin{center}
\includegraphics[width=0.95\linewidth]{EmailVolumeB1LibyaBuildup}
\caption{Feb. 25 -- March 15, 2011:  Selected Emails are those that are B1 redacted}
\label{fig:EmailVolumeB1LibyaBuildup}
\end{center}
\end{figure}
The left most spike in \autoref{fig:EmailVolumeB1LibyaBuildup}  is the rightmost spike of the pair of approximately 80 emails from \autoref{fig:early2011B1}.   
Of course all other displays have been updating every time the time slider is adjusted.

Some sense of the content of the emails can be had from the updated term frequency displays shown in \autoref{fig:TermFreqB1LibyaBuildup}.
\begin{figure}[h]
\begin{center}
\begin{tabular}{ccc}
\includegraphics[width=0.4\linewidth]{TFIDFB1LibyaBuildup} &
$~~~~$ &
\includegraphics[width=0.4\linewidth]{TermFreqB1LibyaBuildup} 
\end{tabular}
\caption{Frequent terms: B1 only emails, Feb. 25 -- March 18, 2011}
\label{fig:TermFreqB1LibyaBuildup}
\end{center}
\end{figure}
Not surprisingly,  Libya and its leader figure prominently in this restricted set of emails.  
Perhaps more surprising is a term like ``windrush''.  A little investigation shows that this is a reference to ``Windrush Ventures'' (e.g. see  \cite{windrushTelegraph, windrushGuardian}),  a company owned by former U.K. Prime Minister Tony Blair .  ``Windrush'' appears at the end of emails as part of Mr. Blair's electronic signature.   
Mr. Blair is part of this B1 selected correspondence.

Other restrictions, as measured by the FOIA exemption codes, can be seen in \autoref{fig:FOIAB12011LibyaBuildup}.
\begin{figure}[h]
\begin{center}
\includegraphics[width=0.95\linewidth]{FOIAB12011LibyaBuildup}
\caption{FOIA code distribution: All  B1 redacted, Feb. 25 -- March 18, 2011}
\label{fig:FOIAB12011LibyaBuildup}
\end{center}
\end{figure}
There appear to be about 100 emails in the selected set.  In addition to having B1 redactions, many of them also have B5 (inter- or intra-agency) and B6 (personal) redactions.

We might now consider the composition of the inner circle for this set of emails.  \autoref{fig:InnerCircle2011B1LibyaBuildup} 
\begin{figure}[h]
\begin{center}
\includegraphics[width=0.95\linewidth]{InnerCircle2011B1LibyaBuildup}
\caption{Inner Circle: All  B1 redacted, Jan. 1 -- Apr 15, 2011}
\label{fig:InnerCircle2011B1LibyaBuildup}
\end{center}
\end{figure}
shows the set of correspondents.  Not surprisingly, Jake Sullivan is the principal correspondent.  Huma Abedin is next, though she is using two emails, one a government account the other a personal \texttt{clintonemail.com} account.   Sidney Blumenthal, a non-government employee, is in this inner circle of B1 exempted emails.  So too is former U.K. Prime Minister Tony Blair; this is the unidentifiable domain account \texttt{aclb}.
It should be noted that there could be other accounts which do not appear in the inner circle because their emails were not present in the email record.  See \autoref{sect:data:metadata}.

We could continue to drill down into this information, checking the (unredacted) contents of the emails themselves from either the Wikileaks or State Department sources.  We might also more closely connect email dates and correspondents with world events.   Instead, we now turn our attention to another feature which appears in \autoref{fig:lastTwoYears}.

\subsection{An email gap?}
\label{sect:emailGap}
Towards the very right of \autoref{fig:lastTwoYears} there appears to be an anomaly in both time series.  There is a noticeable flat area at zero in both series.

If we move the leftmost time sliders to the June 1, 2012, and the rightmost to the end, February 1, 2013, we have the email volumes for the last 8 months of the archive.  The resulting series are shown in \autoref{fig:EmailVolumeB1Jun1On}.
\begin{figure}[h]
\begin{center}
\includegraphics[width=0.95\linewidth]{EmailVolumeB1Jun1On}
\caption{June 1, 2012 -- February 1, 2013:  Selected Emails are those that are B1 redacted}
\label{fig:EmailVolumeB1Jun1On}
\end{center}
\end{figure}
There appears to be no emails for a fairly large chunk of time.  In fact, there are few or no emails in the archive from October 30 to November 9, 2012.  
This seems curious, given that the archive is supposed to be complete in Secretary Clinton's work-related email during this time period.

The block of no email appears as an even more unusual pattern in \autoref{fig:teaser}(a).  There the block is conspicuous by the absence amongst the emails from September 11 to November 23, 2012.  The email on which the displays of \autoref{fig:teaser} are based, is no longer filtered by FOIA code.  It contains all email in the archive in this time period.  Direct examination of the archives at Wikileaks and the State Department show no mail in this empty block beyond that seen in \autoref{fig:teaser}(a).
